Tidy data structure to support exploration and modeling of temporal data

### 2018-12-08 main.Rmd
added paper structure in intro

Suppose that temporal data consists of $N$ subjects/observational units. Given subject $i$ $(i = 1, 2, \dots, N)$, $X_{ijt}$ denotes the $j$th measurement $(j = 1, 2, \dots, p)$ indexed at a unique sequence of time points $1 \le t \le T_{i}$. Temporal data arrives in many possible formats, so does time. An overview of data science workflow is given in Figure \ref{fig:pipeline-ann}. The divide between data that feeds into the beginning of analysis and the current time series standard that fits into modelling is significant, making data preparation for modeling and forecasting painful. For example, data can be recorded at various time resolutions (hours, minutes, and seconds), and they are typically associated with different time zones with adjustments like summer time. It could be irregularly recorded, which is particularly true with longitudinal measurements like patient visits to a doctor's office. Temporal data also often contains rich information: multiple observational units of different time lengths, multiple and heterogeneous measured variables, multiple grouping factors.

@@

The rest of the paper is structured as follows.

@@

Longitudinal data representation has been implemented. The underpinning data structure is a two-dimensional column-homogeneous array, as other tabular data. Specifying a longitudinal data set from a generic array needs explicitly declaring time and individuals (or panel variable in Stata's `tsset` command). Therefore, each row of the data can be identified by a specific individual and time point. Individuals, however, can be only declared through a single variable, not multiple.

@@

There has been a long history of pipeline discussions and implementation centering around data in various aspects. A data pipeline describes the general flow of data through an analysis, and can generally assist in conceptualising the process as it might be applied to a variety of problems. The Extract, Transform, and Load (ETL), from the data warehousing literature, outlines the workflow to prepare data for analysis, dates back to (XXX Caserta, Joe, 1965? REFS needed) can be considered a data pipeline. Building a data pipeline can be technically difficult, to make it sufficiently general for various data, with many implementation decisions on the interface, input and output objects and functionality. It is useful to articulate the data pipeline induced by new data tools.

@@

Modeling is crucial to explanatory and predictive analytics, but often imposes stricter assumptions on tsibble data. The verbs listed in Table \ref{tab:??} ease the transition to a tsibble that suits modeling. A tidy forecasting framework built on top of tsibble is under development, which aims at promoting transparent forecasting practices and concise model representation. A tsibble usually contains multiple time series. Batch forecasting will be enabled if a univariate model, such as ARIMA and Exponential Smoothing, is applied to each time series independently. This yields a "mable" (short for model table), where each model only tags to each "key" value in tsibble to avoid expensive data copying and reduce model storage. The mable is further supplied to forecasting methods, to produce a tsibble in which each "key" along with its future time holds predictions. It also underlines the advantage of tsibble's "key" in linking between data inputs, models and forecasts. Advanced forecasting techniques, such as vector autocorrelation, hierarchical reconciliation, and ensembles, can be developed in alike spirit. The modeling module will be fulfilled and integrated eventually.

@@

# Software structure and design decisions

@@

# Case studies

@@

# Conclusion and future work

>>>>

Suppose that temporal data consists of $N$ subjects/observational units. Given subject $i$ $(i = 1, 2, \dots, N)$, $X_{ijt}$ denotes the $j$th measurement $(j = 1, 2, \dots, p)$ indexed at a unique sequence of time points $1 \le t \le T_{i}$. Temporal data arrives in many possible formats, so does time. For example, time can carry with various time resolutions (hours, minutes, and seconds), and they are typically associated with different time zones with adjustments like summer time. It could be irregularly recorded, which is particularly true with longitudinal measurements like patient visits to a doctor's office. Temporal data also often contains rich information: multiple observational units of different time lengths, multiple and heterogeneous measured variables, multiple grouping factors. An overview of data science workflow is given in Figure \ref{fig:pipeline-ann}. The divide between data that feeds into the beginning of analysis and the current time series standard that fits into modelling is significant, making data preparation for modeling and forecasting painful.

@@

The rest of the paper is structured as follows. Section \ref{sec:str} reviews temporal data structures corresponding to time series and longitudinal analysis, and discusses "tidy data" and the grammar of data manipulation. Section \ref{sec:semantics} proposes contextual semantics for temporal data, built on top of tidy data. The concept of data pipelines with respect to the time domain will be discussed in depth in Section \ref{sec:pipeline}, followed by the design choices of software structure in Section \ref{sec:software}. Section \ref{sec:cases} presents two case studies for temporal data exploration using the newly implemented infrastructure. Section \ref{sec:conclusion} concludes the paper and discusses the future work.

@@

@@

There has been a long history of pipeline discussions and implementation centering around data in various aspects. A data pipeline describes the general flow of data through an analysis, and can generally assist in conceptualising the process as it might be applied to a variety of problems. The Extract, Transform, and Load (ETL), from the data warehousing literature, outlines the workflow to prepare data for analysis, dates back to @kimball2011data can be considered a data pipeline. Building a data pipeline can be technically difficult, to make it sufficiently general for various data, with many implementation decisions on the interface, input and output objects and functionality. It is useful to articulate the data pipeline induced by new data tools.

@@

Modeling is crucial to explanatory and predictive analytics, but often imposes stricter assumptions on tsibble data. The verbs listed in Table \ref{tab:verb-table} ease the transition to a tsibble that suits modeling. A tidy forecasting framework built on top of tsibble is under development, which aims at promoting transparent forecasting practices and concise model representation. A tsibble usually contains multiple time series. Batch forecasting will be enabled if a univariate model, such as ARIMA and Exponential Smoothing, is applied to each time series independently. This yields a "mable" (short for model table), where each model only tags to each "key" value in tsibble to avoid expensive data copying and reduce model storage. The mable is further supplied to forecasting methods, to produce a tsibble in which each "key" along with its future time holds predictions. It also underlines the advantage of tsibble's "key" in linking between data inputs, models and forecasts. Advanced forecasting techniques, such as vector autocorrelation, hierarchical reconciliation, and ensembles, can be developed in alike spirit. The modeling module will be fulfilled and integrated eventually.

@@

# Software structure and design decisions {#sec:software}

@@

# Case studies {#sec:cases}

@@

# Conclusion and future work {#sec:conclusion}
